---
title: "Final Project"
output: html_document
---

##Environment and Meteorology
#Patrick Zivkovic

#Final Project

For this project I chose to take a look on the NAOs effect on precipitation in my home country Austria. It being deeper inside the continent, I do not expect such big effects as we have seen in the cases of Scandinavia or the Iberian penninsula. 
The Summer-NAO (SNAO) on the other hand, the summer counterpart to the NAO, received pretty low amount of attention. Hurrell et al. (2003, 2009) review papers do not say a lot about the SNAO expect that its amplitude and spatial extent are smaller than the NAO and that its effects on precipitation, temperature and cloudiness are shifted northeastward. Bladé, Liebmann and Fortuny (2011) for example provide the information that the SNAO has got some impact on the precipitation in the mediterreanean and Balkan region. This understanding lead me to the idea to test both, the NAO and the SNAOs, effects on the precipitation in two climatically seperable parts of Austria: the Alp region and the eastern, more flat, region.

First we load the packages provided by the climate4R work-group.
```{r}
library(loadeR)
library(transformeR)
library(visualizeR)
library(downscaleR)
library(climate4R.climdex)
```

```{r}
loginUDG('pzi365', '**********')
```

```{r}
lonLim <- c(-90,40)
latLim <- c(20,80)
```

We take the Reanalysis data from 1951-2010 to calculate the NAO index futher down the line.
```{r, eval}
ncep.url <- "http://meteo.unican.es/tds5/dodsC/ncepReanalysis1/ncepReanalysis1_4xDaily.ncml"
```

We load the data considering the variable "sea-level pressure(slp)" with which we are able to calculate the NAO and take a closer look at the winter months (12,1,2). Because this data provides slp at a rate of 6 hours we need to aggregate the data in the way to get first daily data (using the mean) and then monthly data in the next chunk of code even annual data (again using the mean for the aggregation).
```{r,eval=FALSE}
ncep.psl <- loadGridData(dataset = ncep.url,
                         var = "slp",
                         lonLim = lonLim,
                         latLim = latLim,
                         season = c(12,1,2),
                         years = 1951:2010,
                         time = "DD",
                         aggr.d = "mean",
                         aggr.m = "mean")
```

```{r,eval=FALSE}
ncep.psl <- aggregateGrid(grid = ncep.psl, aggr.y = list(FUN = "mean"))
```

To use the principal component analyis effectively (means to be able to take the first component as the one we are interested in and not the second one) we need to calculate the anomaly of the mean of the aggregated slp data.
```{r,message=FALSE}
psl.anom <- scaleGrid(grid = ncep.psl)
```

Using the principal component analysis and plotting the anomaly we see the expected output of the slp pressure being lower in the north atlantic and higher at the latitude levels of central Europe and northern Africa.
```{r}
pca1 <- prinComp(psl.anom, n.eofs = 1)
plotEOF(pca1, var = "psl", n.eofs = 1,
        backdrop.theme = "countries",
        main = "Primera EOF de la anomalía de la SLP")
```
Looking at the now calculated NAO index based on the principal component analysis we can plot the years combined with the index of the corresponding year the generally positive NAO index from 1950-1980 and then the generally negative NAO index until 2010 where we got a peak of observed NAO index values.
```{r}
nao <- PC2grid(pca1, scale = TRUE, opp = TRUE)
nao.index.ncep <- nao[["Data"]][1,,1,1]
years <- getYearsAsINDEX(ncep.psl)
plot(years, nao.index.ncep, ty = 'l', ylab = "NAO Index", xlab = "year")
grid()
pos <- which(nao.index.ncep > 0) ## Index of positive NAO years
neg <- setdiff(1:length(nao.index.ncep), pos) ## Index of negative NAO years
points(years[pos], nao.index.ncep[pos], pch = 19, col = "red")
points(years[neg], nao.index.ncep[neg], pch = 19, col = "blue")
abline(h = 0, lty = 3)
title(main = "PC-based NAO Index ")
```


To analyze the effect of the NAO on the precipitation we load an EOBS dataset of the same period of time taking again just the winter season (12,1,2).
```{r,eval=FALSE}
eobs.url <- "http://opendap.knmi.nl/knmi/thredds/dodsC/e-obs_0.50regular/rr_0.50deg_reg_v17.0.nc"
```

```{r,eval=FALSE}
eobs.precip <- loadGridData(dataset = eobs.url,
                            var = "rr",
                            lonLim = c(-10,20),
                            latLim = c(35,70),
                            season = c(12,1,2),
                            years = 1951:2010,
                            aggr.m = "sum")
```

Instead of the mean function we use the sum function to aggregate the data of precipitation, because we are interested in the total amount of precipitation over a year.
```{r,eval=FALSE}
eobs.precip <- aggregateGrid(eobs.precip, aggr.y = list(FUN = "sum"))
```

Plotting it we can see that Austria generally does not show a lot of precipitation over the winter months.
```{r}
spatialPlot(climatology(eobs.precip),
            backdrop.theme = "countries",
            main = "Precipitación media de invierno (1951-2010)",
            scales=list(draw=TRUE))
```

Finding out the coordinated of the grid of Austria I painted the rectangle encompassing the two climatically seperable Austrian regions (the Alpine and the Eastern one) for further analysis.
```{r}
east_austria <- map.lines(lonLim = c(14.3,17.2), latLim = c(46.2,49.2), col = "green", lwd = 2)
alp_austria <- map.lines(lonLim = c(9,14), latLim = c(46,47.8), col = "blue", lwd = 2)
spatialPlot(climatology(eobs.precip),
            backdrop.theme = "countries",
            sp.layout = list(east_austria, alp_austria))
```

To be able to compare the NAO with the precipitation we need to also calculate the precipitation anomaly over Europe.
```{r}
eobs.precip.anom <- scaleGrid(eobs.precip)
```

Being interested only in Austria we can just select the certain regions using the function 
subsetGrid.
```{r}
eobs.tp.east_austria <- subsetGrid(eobs.precip.anom, lonLim = c(14.3,17.2), latLim = c(46.2,49.2))
eobs.tp.alp_austria <- subsetGrid(eobs.precip.anom, lonLim = c(9,14), latLim = c(46,47.8))
```

```{r}
east_austria.tp <- aggregateGrid(eobs.tp.east_austria,
                           aggr.lon = list(FUN = "mean", na.rm = TRUE),
                           aggr.lat = list(FUN = "mean", na.rm = TRUE))
alp_austria.tp <- aggregateGrid(eobs.tp.alp_austria,
                           aggr.lon = list(FUN = "mean", na.rm = TRUE),
                           aggr.lat = list(FUN = "mean", na.rm = TRUE))
```

Plotting the anomalies against each other we do really see some consistent concurrencies.
```{r}
plot(1951:2010, nao.index.ncep, ty = "o", xlab = "year", ylab = "Standardized Value")
lines(1951:2010, scale(east_austria.tp$Data), col = "red")
grid()
legend("bottomleft", c("NAO", "Winter Precip Anomaly"), lty = 1, col = c("black", "red"), bty = "n")
title(main = "Domain of East Austria")
mtext(paste("Correlation =", round(cor(nao.index.ncep, east_austria.tp$Data),3)))
```

Finally calculating the correlation coefficient by Pearson we see a value of approximately 0.38 with a significant t-test (p = 0.0025) indicating a significant connection between these two variables.
```{r}
cor.test(nao.index.ncep, east_austria.tp$Data)
```

```{r}
plot(1951:2010, nao.index.ncep, ty = "o", xlab = "year", ylab = "Standardized Value")
lines(1951:2010, scale(alp_austria.tp$Data), col = "red")
grid()
legend("bottomleft", c("NAO", "Winter Precip Anomaly"), lty = 1, col = c("black", "red"), bty = "n")
title(main = "Domain of Alpine Austria")
mtext(paste("Correlation =", round(cor(nao.index.ncep, alp_austria.tp$Data),3)))
```

A similiar result to the domain of East Austria we got a correlation of 0.37 for the alp region with p = 0.0034.
```{r}
cor.test(nao.index.ncep, alp_austria.tp$Data)
```

#################################################################################
After establishing that there is not a sufficiently convincing connection between the NAO index and the precipitation in Austria, I asked myself the question if the summer North Atlantic Oscillation (SNAO) maybe has an impact on the same. This SNAO is defined as the first empirical orthogonal function (EOF) of observed summertime North Atlantic pressure at sea level just like its winter counterpart for the wintertime respectively.
So the following code is basically the same with changes in the data loading (taking the summer months instead of the winter ones).

```{r,eval=FALSE}
ncep.psl_summer <- loadGridData(dataset = ncep.url,
                         var = "slp",
                         lonLim = lonLim,
                         latLim = latLim,
                         season = c(6,7,8),
                         years = 1951:2010,
                         time = "DD",
                         aggr.d = "mean",
                         aggr.m = "mean")
```

```{r,eval=FALSE}
ncep.psl_summer <- aggregateGrid(grid = ncep.psl_summer, aggr.y = list(FUN = "mean"))
```

```{r,message=FALSE}
psl.anom_summer <- scaleGrid(grid = ncep.psl_summer)
```

```{r}
pca1_summer <- prinComp(psl.anom_summer, n.eofs = 1)
plotEOF(pca1_summer, var = "psl", n.eofs = 1,
        backdrop.theme = "countries",
        main = "Primera EOF de la anomalía de la SLP")
```

```{r}
snao <- PC2grid(pca1_summer, scale = TRUE, opp = TRUE)
snao.index.ncep_summer <- snao[["Data"]][1,,1,1]
years <- getYearsAsINDEX(ncep.psl_summer)
plot(years, snao.index.ncep_summer, ty = 'l', ylab = "SNAO Index", xlab = "year")
grid()
pos <- which(snao.index.ncep_summer > 0) ## Index of positive SNAO years
neg <- setdiff(1:length(snao.index.ncep_summer), pos) ## Index of negative SNAO years
points(years[pos], snao.index.ncep_summer[pos], pch = 19, col = "red")
points(years[neg], snao.index.ncep_summer[neg], pch = 19, col = "blue")
abline(h = 0, lty = 3)
title(main = "PC-based SNAO Index ")
```


To analyze the effect of the SNAO on the precipitation we load an EOBS dataset of the same period of time taking again just the summer season (6,7,8).

```{r,eval=FALSE}
eobs.precip_summer <- loadGridData(dataset = eobs.url,
                            var = "rr",
                            lonLim = c(-10,20),
                            latLim = c(35,70),
                            season = c(6,7,8),
                            years = 1951:2010,
                            aggr.m = "sum")
```

```{r,eval=FALSE}
eobs.precip_summer <- aggregateGrid(eobs.precip_summer, aggr.y = list(FUN = "sum"))
```

Plotting it we can see that large parts of Austria generally show a lot of precipitation over the summer months.
```{r}
spatialPlot(climatology(eobs.precip_summer),
            backdrop.theme = "countries",
            main = "Precipitación media de verano (1951-2010)")
```

To be able to compare the SNAO with the precipitation we need to also calculate the precipitation anomaly over Europe.
```{r}
eobs.precip.anom_summer <- scaleGrid(eobs.precip_summer)
```

```{r}
eobs.tp.east_austria_summer <- subsetGrid(eobs.precip.anom_summer, lonLim = c(14.3,17.2), latLim = c(46.2,49.2))
eobs.tp.alp_austria_summer <- subsetGrid(eobs.precip.anom_summer, lonLim = c(9,14), latLim = c(46,47.8))
```

```{r}
east_austria.tp_summer <- aggregateGrid(eobs.tp.east_austria_summer,
                           aggr.lon = list(FUN = "mean", na.rm = TRUE),
                           aggr.lat = list(FUN = "mean", na.rm = TRUE))
alp_austria.tp_summer <- aggregateGrid(eobs.tp.alp_austria_summer,
                           aggr.lon = list(FUN = "mean", na.rm = TRUE),
                           aggr.lat = list(FUN = "mean", na.rm = TRUE))
```

```{r}
plot(1951:2010, snao.index.ncep_summer, ty = "o", xlab = "year", ylab = "Standardized Value")
lines(1951:2010, scale(east_austria.tp_summer$Data), col = "red")
grid()
legend("bottomleft", c("SNAO", "Summer Precip Anomaly"), lty = 1, col = c("black", "red"), bty = "n")
title(main = "Domain of East Austria")
mtext(paste("Correlation =", round(cor(snao.index.ncep_summer, east_austria.tp_summer$Data),3)))
```


Finally calculating the correlation coefficient by Pearson we see a value of approximately 0.365 with a significant t-test (p = 0.004) indicating a significant connection between these two variables.
```{r}
cor.test(nao.index.ncep_summer, east_austria.tp_summer$Data)
```

```{r}
plot(1951:2010, snao.index.ncep_summer, ty = "o", xlab = "year", ylab = "Standardized Value")
lines(1951:2010, scale(alp_austria.tp_summer$Data), col = "red")
grid()
legend("bottomleft", c("SNAO", "Summer Precip Anomaly"), lty = 1, col = c("black", "red"), bty = "n")
title(main = "Domain of Alp Austria")
mtext(paste("Correlation =", round(cor(snao.index.ncep_summer, alp_austria.tp_summer$Data),3)))
```

For the alpine region we get a bigger correlation coefficient, namely 0.452 with p = 0.0003 indicating a bigger relationship between the precipitation of the summer months in the alps and the SNAO. 
```{r}
cor.test(snao.index.ncep_summer, alp_austria.tp_summer$Data)
```

Taking only the positive or the negative SNAO indices did not improve the correlation, so there seem to be no specific correlations with the two extremes.

In conclusion, the NAO index seem to moderately affect the precipitation in both specified Austrian regions over the winter months (with correlations of around 0.38). But taking a look at the SNAO in connection with the precipitation in the summer months, there seems to be a stronger connection, especially in the Alpine region of Austria (namely r = 0.45), further advancing the idea that the NAO/SNAO take a huge role in the states of climate across Europe.

Literature:

Bladé, I., Liebmann, B., Fortuny, D., & van Oldenborgh, G. J. (2012). Observed and simulated impacts of the summer NAO in Europe: implications for projected drying in the Mediterranean region. Climate dynamics, 39(3-4), 709-727.

Hurrell JW, Kushnir Y, Ottersen G, Visbeck M (2003) An overview
of the North Atlantic Oscillation. The North Atlantic Oscillation:
climatic significance and environmental impact. Geophys Monogr.
Am Geophys Union 134:1–35

Hurrell JW, Deser C (2009) North Atlantic climate variability: the
role of the North Atlantic Oscillation. J Mar Syst 78(1):28–41.
doi:10.1016/j.jmarsys.2008.11.026


